---
title: "Project"
author: "Sarang Pande"
date: "4/12/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Libraries

```{r}
library(dplyr)
library(leaps)
library(tidyverse)
library(cowplot)
library(GGally)
library(Metrics)
library(glmnet)
library(pls)
library(caret)
library(nnet)
library(tree)
library(gridExtra)
```

# Data Gathering

```{r}
loan_data <- read.csv('loan.csv')
```

# Data Preprocessing

Lets extract all the borrower characteristic available before he applies for a loan. 

```{r}
column_vector <- c('loan_amnt', 'term', 'int_rate', 'emp_length', 'home_ownership', 'annual_inc', 'verification_status', 'purpose', 'addr_state', 'dti', 'open_acc', 'total_acc', 'grade', 'sub_grade', 'delinq_2yrs', 'inq_last_6mths', 'pub_rec', 'revol_bal', 'revol_util')
loan_data_filtered <- loan_data %>% select(all_of(column_vector))
```

Lets drop rows missing any of the data.

```{r}
loan_data_filtered <- drop_na(loan_data_filtered)
```

```{r}
summary(loan_data_filtered)
```

```{r}
a <- loan_data_filtered %>% select(home_ownership, annual_inc, dti, revol_util)
plot1 <- ggplot(a) + geom_histogram(aes(annual_inc), bins = 30, color = 'red') + theme(axis.text.x = element_text(angle = 45, hjust = 1))
plot2 <- ggplot(a) + geom_bar(aes(home_ownership)) + theme(axis.text.x = element_text(angle = 45, hjust = 1))
plot3 <- ggplot(a) + geom_histogram(aes(dti), bins = 30) + theme(axis.text.x = element_text(angle = 45, hjust = 1))
plot4 <- ggplot(a) + geom_histogram(aes(revol_util), bins = 30) + theme(axis.text.x = element_text(angle = 45, hjust = 1))
grid.arrange(plot1, plot2, plot3, plot4, ncol=2)
```
# Data Transformation

```{r}
table(loan_data_filtered$emp_length)
loan_data_filtered <- loan_data_filtered %>% filter(emp_length != 'n/a')
```

```{r}
loan_data_filtered <- loan_data_filtered %>% mutate(years = ifelse(emp_length ==  "< 1 year", 0, ifelse(emp_length ==  "1 year", 1, ifelse(emp_length ==  "2 years", 2, ifelse(emp_length ==  "3 years", 3, ifelse(emp_length ==  "4 years", 4, ifelse(emp_length ==  "5 years", 5, ifelse(emp_length ==  "6 years", 6, ifelse(emp_length ==  "7 years", 7, ifelse(emp_length ==  "8 years", 8, ifelse(emp_length ==  "9 years", 9, 10)))))))))))
loan_data_filtered <- loan_data_filtered %>% select(-emp_length)
loan_data_filtered$years <- as.factor(loan_data_filtered$years)
```

```{r}
table(loan_data_filtered$home_ownership)
loan_data_filtered$home_ownership <- as.character(loan_data_filtered$home_ownership)
loan_data_filtered <- loan_data_filtered %>% filter(home_ownership %in% c('MORTGAGE', 'OWN', 'RENT'))
loan_data_filtered$home_ownership <- as.factor(loan_data_filtered$home_ownership)
```

```{r}
boxplot(loan_data_filtered$annual_inc)
hist(loan_data_filtered$annual_inc, breaks = 100)
```

```{r}
outliers <- boxplot(loan_data_filtered$annual_inc, plot = FALSE)$out
loan_data_filtered <- loan_data_filtered[-which(loan_data_filtered$annual_inc %in% outliers),]
```

```{r}
class(loan_data_filtered$purpose)
loan_data_filtered$purpose <- as.character(loan_data_filtered$purpose)
```

```{r}
table(loan_data_filtered$purpose)
loan_data_filtered <- loan_data_filtered %>% mutate(purpose = ifelse(purpose %in% c('educational', 'house', 'vacation', 'wedding', 'renewable_energy', 'moving'), 'other', purpose))
loan_data_filtered$purpose <- as.factor(loan_data_filtered$purpose)
```

```{r}
boxplot(loan_data_filtered$dti)
hist(loan_data_filtered$dti, breaks = 100)
```

```{r}
outliers <- boxplot(loan_data_filtered$dti, plot = FALSE)$out
loan_data_filtered <- loan_data_filtered[-which(loan_data_filtered$dti %in% outliers),]
loan_data_filtered <- loan_data_filtered %>% filter(dti >= 0)
```

```{r}
boxplot(loan_data_filtered$open_acc)
hist(loan_data_filtered$open_acc, breaks = 100)
```

```{r}
outliers <- boxplot(loan_data_filtered$open_acc, plot = FALSE)$out
#loan_data_filtered <- loan_data_filtered[-which(loan_data_filtered$open_acc %in% outliers),]
```

```{r}
boxplot(loan_data_filtered$total_acc)
hist(loan_data_filtered$total_acc, breaks = 100)
```

```{r}
outliers <- boxplot(loan_data_filtered$total_acc, plot = FALSE)$out
#loan_data_filtered <- loan_data_filtered[-which(loan_data_filtered$total_acc %in% outliers),]
```

```{r}
boxplot(loan_data_filtered$revol_bal)
hist(loan_data_filtered$revol_bal, breaks = 100)
```

```{r}
outliers <- boxplot(loan_data_filtered$revol_bal, plot = FALSE)$out
#loan_data_filtered <- loan_data_filtered[-which(loan_data_filtered$revol_bal %in% outliers),]
```

```{r}
boxplot(loan_data_filtered$revol_util)
hist(loan_data_filtered$revol_util, breaks = 100)
```

```{r}
outliers <- boxplot(loan_data_filtered$revol_util, plot = FALSE)$out
loan_data_filtered <- loan_data_filtered[-which(loan_data_filtered$revol_util %in% outliers),]
```

```{r}
#loan_data_filtered$pub_rec <- as.factor(loan_data_filtered$pub_rec)
#loan_data_filtered$inq_last_6mths <- as.factor(loan_data_filtered$inq_last_6mths)
#loan_data_filtered$delinq_2yrs <- as.factor(loan_data_filtered$delinq_2yrs)
loan_data_filtered <- loan_data_filtered %>% select(-addr_state)
```

```{r}
normalize <- function(x)
{
    return((x- min(x)) /(max(x)-min(x)))
}

loan_data_filtered$loan_amnt <- sapply(list(loan_data_filtered$loan_amnt), normalize)
loan_data_filtered$annual_inc <- sapply(list(loan_data_filtered$annual_inc), normalize)
loan_data_filtered$revol_bal <- sapply(list(loan_data_filtered$revol_bal), normalize)
```


# Modelling

## Forward Selection

```{r}
regfit.fwd = regsubsets(int_rate ~ ., data = loan_data_filtered, really.big = TRUE, method ="forward", nvmax = 30)
```

```{r}
regfit.fwd.summary <- summary(regfit.fwd)
par(mfrow=c(2, 2))
plot(regfit.fwd.summary$rss ,xlab="Number of Variables ",ylab="RSS",type="l") 
plot(regfit.fwd.summary$adjr2 ,xlab="Number of Variables ", ylab="Adjusted RSq",type="l")
plot(regfit.fwd.summary$cp ,xlab="Number of Variables ",ylab="Cp", type='l')
plot(regfit.fwd.summary$bic ,xlab="Number of Variables ",ylab="BIC",type='l')
```

```{r}
coef(regfit.fwd, 6)
```

```{r}
plot(regfit.fwd,scale="r2")
plot(regfit.fwd,scale="adjr2")
plot(regfit.fwd,scale="Cp")
plot(regfit.fwd,scale="bic")
```

```{r}
temp <- loan_data_filtered %>% select(-c('grade','sub_grade'))
regfit.fwd.v2 = regsubsets(int_rate ~ ., data = temp, really.big = TRUE, method ="forward", nvmax = 40)
```

```{r}
plot(regfit.fwd.summary$adjr2 ,xlab="Number of Variables ", ylab="Adjusted RSq",type="l", col='red')
lines(regfit.fwd.v2.summary$adjr2 ,xlab="Number of Variables ", ylab="Adjusted RSq",type="l", col = 'green')
legend("bottomright", legend=c("With Grades", "Without Grades"),col=c("red", "green"), lty=1:2, cex=0.8)
title('Forward selection with and without grades')
```

```{r}
regfit.fwd.v2.summary <- summary(regfit.fwd.v2)
par(mfrow=c(2, 2))
plot(regfit.fwd.v2.summary$rss ,xlab="Number of Variables ",ylab="RSS",type="l") 

plot(regfit.fwd.v2.summary$cp ,xlab="Number of Variables ",ylab="Cp", type='l')
plot(regfit.fwd.v2.summary$bic ,xlab="Number of Variables ",ylab="BIC",type='l')
```

```{r}

```

```{r}
coef(regfit.fwd.v2, 16)
```

```{r}
plot(regfit.fwd.v2,scale="r2")
plot(regfit.fwd.v2,scale="adjr2")
plot(regfit.fwd.v2,scale="Cp")
plot(regfit.fwd.v2,scale="bic")
```

## Lasso Regression

```{r}
y <- loan_data_filtered$int_rate
x <- loan_data_filtered %>% select(-int_rate) %>% data.matrix()
lambda_seq <- 10^seq(2, -2, by = -.1)

# Splitting the data into test and train
set.seed(7)
train = sample(1:nrow(x), nrow(x)/2)
x_test = (-train)
y_test = y[-train]

cv_output <- cv.glmnet(x[train,], y[train], alpha = 1, lambda = lambda_seq)

# identifying best lamda
best_lam <- cv_output$lambda.min

#lambdas <- 10^seq(3, -2, by = -.1)
#cv_fit <- cv.glmnet(x, y, alpha = 0, lambda = lambdas)
#plot(cv_fit)
```

```{r}
lasso_best <- glmnet(x[train,], y[train], alpha = 1, lambda = best_lam)
```

```{r}
coef(lasso_best)
```

## Principal Component Regression

```{r}
set.seed(2)
pcr.fit=pcr(int_rate ~ ., data=loan_data_filtered , scale=TRUE , validation ="CV")
```

```{r}
b <- pcr.fit
#summary(pcr.fit)
validationplot(b, val.type="MSEP")
```

```{r}
set.seed(2)
pcr.temp.fit=pcr(int_rate ~ ., data=temp , scale=TRUE , validation ="CV")
```

```{r}
#summary(pcr.temp.fit)
plot1 <- validationplot(pcr.temp.fit, val.type="MSEP")
plot2 <- validationplot(b, val.type="MSEP")
grid.arrange(plot1, plot2, ncol = 2)
```

## Partial Least Squares

```{r}
pls.fit=plsr(int_rate ~ ., data=loan_data_filtered , scale=TRUE , validation ="CV")
pls.temp.fit=pcr(int_rate ~ ., data=temp , scale=TRUE , validation ="CV")
```

```{r}
validationplot(pls.fit, val.type="R2")
validationplot(pls.temp.fit, val.type="R2")
```

## Logistic Regression

```{r}
new_temp <- loan_data_filtered %>% select(-c(sub_grade, int_rate, years))
table(new_temp$grade)
```


## Upsampling

```{r}
new_temp$grade <- as.character(new_temp$grade)
new_temp_1 <- new_temp %>% filter(grade != 'G')
new_temp_1$grade <- as.factor(new_temp_1$grade)
x_1 <- new_temp_1 %>% select(-grade)
data_1 <- downSample(x = x_1, y = new_temp_1$grade)

new_temp_2 <- new_temp %>% filter(grade %in% c('F', 'G'))
new_temp_2$grade <- as.factor(new_temp_2$grade)
x_2 <- new_temp_2 %>% select(-grade)
data_2 <- upSample(x = x_2, y = new_temp_2$grade)
data_2 <- data_2 %>% filter(Class == 'G')

data <- rbind(data_1, data_2)
```

## Downsampling
```{r}
x <- new_temp %>% select(-grade)
set.seed(234)
data <- downSample(x = x, y = new_temp$grade)
```

## Sampling

```{r}
new_temp <- sample_n(new_temp, 78000)
table(new_temp$grade)
data <- new_temp %>% mutate(Class = grade) %>% select(-grade)
```

```{r}
table(data$Class)
```

```{r}
set.seed(7)
train_indices <- sample(seq_len(nrow(data)), size = 0.75*nrow(data))
train_data <- data[train_indices,]
test_data <- data[-train_indices,]
```

```{r}
glm.fits = multinom(Class ~ ., data=train_data)
summary(glm.fits)
```

```{r}
predicted_class <- predict(glm.fits, test_data)
```

```{r}
write.csv(table(predicted_class, test_data$Class))
```


## Decision Tree

```{r}
tree.fit = tree(Class ~ ., train_data)
summary(tree.fit)
```

## Random Forest

```{r}
tree.fit = tree(Class ~ ., train_data)
summary(tree.fit)
```